Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-23 paper/paper.tex
Revise related work

Autosubst 2 takes our syntax specification in higher-order abstract
syntax and generates Coq syntax specification, renaming and
substitution functions, and lemmas and tactics that allow reasoning
about those functions. The auto-generated files and the library header
files are counted toward the \emph{Library (Autosubst 2)} category.
\scw{this line is no longer in the figure}

@@

\section{Related Work and Discussion}
\label{sec:relatedwork}

@@

\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that establish metatheoretic results strong enough to derive
consistency for dependently typed systems.
\scw{Should give Martin L\"of credit for first consistency proof.}
\scw{What techniques do these three proofs use? How is it different from
your proof? What about Luo, which
introduced a universe hierarchy? What about categories with families, or
other ways of showing consistency?}
Some of these techniques
are still in use in more recent works. For example, the technique for
proving strong normalization by \citet{geuvers1994short} is adapted by
\citet{moon2021graded} to show the same property to a dependently
typed system extended with modalities.

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
uses a Kripke-style logical relation parameterized over an
type-directed equivalence relation satisfying certain
properties. Their logical relation is defined using the
induction-recursion scheme, which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq using a
special way of encoding induction recursion. Their work further
extends the decidability of type conversion result from~\citet{decagda} to the decidability
type checking of a bidrectional type system.

\citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanizes the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
\citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. However, instead of explaining the purely inductive
logical relation as a convoluted workaround to the lack of
induction-recursion in Coq, we give a self-contained explanation of
our logical relation.

Finally, it is worth pointing out that all systems we have discussed
so far builds a relational model for their logical relation. Recall
that our logical relation takes the form $[[InterpR i A S]]$ where
$[[S]]$ is a set of terms. In a relational model, each type is
interpreted as a partial equivalence relation over terms rather than a
set. A relational model is necessary for an extensional type theory
such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
use a relational model either to justify the function $\eta$-law or to
derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
law for function is baked into the untyped reduction
relation and therefore a relational model is not needed.
\scw{would it be difficult to extend your proof to be a PER?}

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We hope features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts can be implemented in a similarly streamlined
fashion and we will leave those extensions as part of our future work.

% To end this section, we
\scw{should this be a new subsection?}
The final question we want to address is: why is our proof,
even with the extension from Section~\ref{sec:extension}, so much
shorter than the proofs from \citet{decagda, nbeincoq,
  martin-lof-a-la-coq}?
First, unlike developments that mechanize the correctness of a clever
type conversion algorithm, we only show the existence of
normal form for open and closed terms and state our properties in
terms of the untyped small-step reduction relation. This removes a lot
of scaffolding in the specification of our type system and our logical
relation.

This contrasts our development with
\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
difficult property to prove due to the complexity of the equality
judgment used to define the logical relation. In our system, the
irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
lemma name} which is
straightforward after we prove the results about parallel
reduction in Section~\ref{sec:spec}. In other words, we keep the
the logical relation itself minimal at the cost of extra lemmas that can be proven
through syntactic means independently from the logical relation.
For example, as commented near the end of Section~\ref{sec:extension},
our normalization property (Corollary~\ref{corollary:exbetanf})
% can induce a decision procedure for type conversion by reducing two
% terms to $\beta\eta$-normal form and check for their syntactic
% equivalence. This decision procedure would be inefficient compared to
% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
% to weak head normal and checks for the equivalence of subterms
% recursively
does not immediately induce an efficient decision procedure, but it is
possible to recover an efficient algorithm by reasoning about untyped
lambda terms.

@@

Another key to our proof is the very early establishment of the confluence
property of the paralell reduction relation
(Lemma~\ref{lemma:pardiamond}). This property is used in
Lemma~\ref{lemma:interppreservation} to show that the interpretation
of a type is preserved under evaluation. That is, if $[[InterpR i A
S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
In a system with typed directed reduction, the confluence result is
harder to establish since confluence requires subject reduction, which
circularly depends on $\Pi$-injectivity, a consequence of
confluence~\citep{siles2012pure}. There are different workarounds to
this problem. \citet{siles2012pure} proposes a syntactic approach by
defining a type system where the confluence result is
directly provable and later show that the system of interest is
equivalent to the system with the confluence property.
\citet{decagda} uses a relational model and the relational counterpart
of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
model is parameterized by a typed indexed equivalence relation and
is reused to prove derive different properties in their
development. Another perhaps simpler approach is to simply
extend the judgmental equality with rules about
$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
reduction to be proven independently from confluence and
$\Pi$-injectivity can later be shown to be admissible. We find the
approach from \citep{weirich:systemd} the most lightweight as it
allows us to derive confluence early on even in systems with typed
conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}

In cases where the judgmental equality does not leverage type information heavily
(e.g. $\eta$-law for the unit type), it might be possible to show that
$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
property holds (we believe this is likely true for both
\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
terms of untyped parallel reduction to show the decidability of type
conversion. Otherwise, to fully model the system, we need to bake
typing information into our logical relation and we need to redefine
our logical relation in Kripke-style so we can relax the scoping
constraint in order to prove adequacy, though a relational model would
still be unnecessary as long as the $\eta$ laws are baked into the
typed directed relation. \scw{not sure I understand this last bit}

Since confluence plays such as key role in our proof, we are uncertain
how to generalize our technique to systems where confluence does not
hold, and we will investigate those scenarios in our future work.
\scw{What systems are you thinking of? Or do you mean more generally
that this technique only works for languages where the definition of
equality is based on joinability of parallel reduction?}

>>>>

Autosubst 2 takes our syntax specification in higher-order abstract
syntax and generates Coq syntax specification, renaming and
substitution functions, and lemmas and tactics that allow reasoning
about those functions. The auto-generated syntax file by Autosubst 2
and library header files are omitted from the figure.

@@

\section{Related Work}
\label{sec:relatedwork}

@@

Proof by logical relations does not come with a precise definition. In
the most general sense, a logical relation can be viewed as a
practical technique that uses a type-indexed relation to assign
meanings to types and strengthen the induction hypothesis for the
property of interest. The original idea of this technique can be
traced back to
\citet{tait1967:reducibility}. \citet{tait1967:reducibility} maps
types to sets of terms satisfying certain properties related to reduction.
The same idea is explained in \citet{girard1989proofs} and extended to
prove the strong normalization property of System F.

Tait's method has been successfully applied to dependently typed
languages.  \citet{Martin-Lof-1973}, \citet{luo1990extended},
\citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that prove strong
normalization for different dependently type systems using Tait's notion of
reducibility.

\scw{Should give Martin L\"of credit for first consistency proof.}
\scw{What techniques do these three proofs use? How is it different from
your proof? What about Luo, which
introduced a universe hierarchy? What about categories with families, or
other ways of showing consistency?}

% Some of these proofs are still reapplied in more recent works. For
% example, the technique for proving strong normalization by
% \citet{geuvers1994short} is adapted by \citet{moon2021graded} to show
% the normalization property for a dependently typed system extended with
% modalities.

The pen and paper representation of the logical relation proofs
can be challenging to adapt to a theorem prover since a lot of details
are hidden behind the concise notations.
As an example, \citet{geuvers1994short} presents the interpretation for types as
an inductively defined total function over the set of syntactically
well-formed types. Despite the notation for the logical relation as a
simply typed function that takes a type and returns some set, the
interpretation function in reality is a dependent function whose
return type depends on the derivation of the well-typedness of its
input. The well-typedness derivation and the proof of the
classification theorem are needed relevantly in the body
of the interpretation function to decide whether an argument of an
application should be erased during interpretation. Due to the
impredicativity of the object language, \citet{geuvers1994short}'s
proof cannot be encoded in Agda, which has a predicative
metatheory. Due to the use of proof-relevant derivations, even in
Coq, a proof assistant that supports impredicativity, one would need
to constantly juggle between the impredicative but irrelevant sort
\texttt{Prop} sort and the predicative but relevant sort
\texttt{Type}.

More recent works such as \citet{Abel12}
and\citet{abel2008betaeta} make their definitions more explicit and
precise and thus easier to encode in proof assistants. Our definition
of logical relation is reminiscent to their definition of a
semantic universe hierarchy, though our logical relation is defined to
be closed under expansion with respect to parallel reduction rather
than weak head reduction. Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use the semantic universe hierachy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly because it is sufficient
for deriving consistency and later normalization.

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facillitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidrectional type system.

\citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanizes the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
\citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. However, instead of explaining the inductive
logical relation as a convoluted workaround to the lack of
induction-recursion in Coq, we give a self-contained explanation of
our logical relation.

Finally, \citet{barras2010sets, Wang2013SemanticsOI} assign
set-theoretic semantics to dependent type theory in Coq. Unlike the
previous mechanization efforts, which primarily focus on predicative
type theory and more direct reducibility models,
\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
$CC^\omega$, a system that incorporates a predicative universe on top
of the impredicative sort in Calculus of Constructions. We choose to
focus on a simple term model so we do not have to take the extra step
of mechanizing mathematical objects such as sets and domains.

\section{Discussion}
\label{sec:discuss}
\yl{I'm not sure how strong of a statement we can make about our proof
technique. We've already demonstrated how to address type-level
dependency when defining a logical relation through a simple
example. Claiming that our proof structure is better seems quite
ambitious, but I think there's a middle point where we claim that
adding moderate features like typed reduction doesn't instantly make
our code size expand all the way from 1000 to 20,000 without saying
the other developments are just verbose for no good reason}
The short consistency proof we
present achieves the goal of ilustrating the technique of proof by
logical relation for dependently typed languages.
In this section, we discuss what attributes to the shortness of our
proof.
% focus on the following question: why is our proof,
% even with its extension to the existence $\beta\eta$-normal form, so much
% shorter than the other mechanized results from \citet{decagda,
% nbeincoq, martin-lof-a-la-coq}?

First, unlike developments that mechanize the correctness of a
type-directed conversion algorithm, we only show the existence of
normal form for open and closed terms and state our properties in
terms of the untyped small-step reduction relation. This removes a lot
of scaffolding related to the specification of the algorithm and the
extra proof obligation that the algorithm is sound and complete with
respect to the declarative specification of the type system.

Second, our logical relation, unlike the ones from \citet{decagda,
  nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
predicates rather than relations. The need for a relational model is
directly related to the metatheoretic property one wants to
prove. \citet{anand2014towards} requires a relational model to capture
the extensionality of their type system, wheras \citet{decagda,
  martin-lof-a-la-coq, nbeincoq} uses a relational model to derive the
injectivity of $\Pi$ types and justify the validity of
$\eta$-conversion among other properties.
Since \lang{} uses an untyped conversion rule, type conversion can be
done through the \emph{normalize-and-compare}
strategy described in \citet{pierce2004advanced}. The decidability of
normalize-and-compare is implied by the existence of
$\beta\eta$-normal form, which follows from our logical
predicate.

In \rref{I-Par, I-Bool}, we close over our semantically valid types
and terms using the non-deterministic parallel reduction relation,
which relates to full reduction, whereas \citet{decagda,nbeincoq} all
refer to the deterministic weak head reduction. The use of a
non-deterministic reduction strategy comes at the cost that we need
confluence to prove the functionality of our logical relation, but
immediately gives us a semantic justification of the conversion rule
(Lemma~\ref{lemma:logrelcoherence}).
% In particular, we obtain the confluence
% result of reduction at a very early stage before we even define our
% logical relation.

With untyped conversion,
we sidestep the relational, Kripke-style logical relation found in
other mechanized proofs. However, our early dependence on confluence
before the fundamental theorem is established can be alarming.
In a system with type-directed reduction,
confluence is not immediately available since it depends on
depends on $\Pi$-injectivity, which is usually only proven after the
fundamental theorem.
Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
problem that allow us to recover confluence property independently
from the logical relation. \citet{siles2012pure} generalizes the
notion of Type Parallel One Step Reduction from \citet{adams2006pure}
to syntactically prove $\Pi$-injectivity for arbitrary Pure Tyep
Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
equational theory, and thus allowing subject reduction to be proven
independently from confluence. Therefore, we believe that even in a
system where type-directed reduction is required (e.g. a system with
the unit $\eta$-law) in the logical relation, the proof can still be carried
out in a structure similar to the one we have presented.

% Finally, it is worth pointing out that all systems we have discussed
% so far builds a relational model for their logical relation.

% Recall
% that our logical relation takes the form $[[InterpR i A S]]$ where
% $[[S]]$ is a set of terms. In a relational model, each type is
% interpreted as a partial equivalence relation over terms rather than a
% set. A relational model is necessary for an extensional type theory
% such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
% use a relational model either to justify the function $\eta$-law or to
% derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
% law for function is baked into the untyped reduction
% relation and therefore a relational model is not needed.
% \scw{would it be difficult to extend your proof to be a PER?}

% The dependently typed systems we have discussed so far also vary
% greatly in expressiveness. For example, the Calculus of
% Constructions~\citep{CoC} and MLTT without universe levels are unable
% to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
% identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
% Among
% the mechanized systems we have discussed so far, only
% \citet{anand2014towards} has a full universe hierarchy.
% Our object language \lang{}, while small, has a decent coverage of
% features commonly seen in dependently typed languages, including
% dependent pattern matching, dependent function types, a full universe
% hierarchy, an intensional identity type. We hope features such as
% type-directed reduction (necessary for unit $\eta$-law) and
% impredicative sorts can be implemented in a similarly streamlined
% fashion and we will leave those extensions as part of our future work.

% To end this section, we
% \scw{should this be a new subsection?}
% The final question we want to address is: why is our proof,
% even with the extension from Section~\ref{sec:extension}, so much
% shorter than the proofs from \citet{decagda, nbeincoq,
%   martin-lof-a-la-coq}?

% This contrasts our development with
% \citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
% difficult property to prove due to the complexity of the equality
% judgment used to define the logical relation. In our system, the
% irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
% direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
% lemma name} which is
% straightforward after we prove the results about parallel
% reduction in Section~\ref{sec:spec}. In other words, we keep the
% the logical relation itself minimal at the cost of extra lemmas that can be proven
% through syntactic means independently from the logical relation.
% For example, as commented near the end of Section~\ref{sec:extension},
% our normalization property (Corollary~\ref{corollary:exbetanf})
% % can induce a decision procedure for type conversion by reducing two
% % terms to $\beta\eta$-normal form and check for their syntactic
% % equivalence. This decision procedure would be inefficient compared to
% % NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
% % to weak head normal and checks for the equivalence of subterms
% % recursively
% does not immediately induce an efficient decision procedure, but it is
% possible to recover an efficient algorithm by reasoning about untyped
% lambda terms.

@@

% Another key to our proof is the very early establishment of the confluence
% property of the paralell reduction relation
% (Lemma~\ref{lemma:pardiamond}). This property is used in
% Lemma~\ref{lemma:interppreservation} to show that the interpretation
% of a type is preserved under evaluation. That is, if $[[InterpR i A S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
% In a system with typed directed reduction, the confluence result is
% harder to establish since confluence requires subject reduction, which
% circularly depends on $\Pi$-injectivity, a consequence of
% confluence~\citep{siles2012pure}. There are different workarounds to
% this problem. \citet{siles2012pure} proposes a syntactic approach by
% defining a type system where the confluence result is
% directly provable and later show that the system of interest is
% equivalent to the system with the confluence property.
% \citet{decagda} uses a relational model and the relational counterpart
% of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
% model is parameterized by a typed indexed equivalence relation and
% is reused to prove derive different properties in their
% development. Another perhaps simpler approach is to simply
% extend the judgmental equality with rules about
% $\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
% reduction to be proven independently from confluence and
% $\Pi$-injectivity can later be shown to be admissible. We find the
% approach from \citep{weirich:systemd} the most lightweight as it
% allows us to derive confluence early on even in systems with typed
% conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}

% In cases where the judgmental equality does not leverage type information heavily
% (e.g. $\eta$-law for the unit type), it might be possible to show that
% $\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
% $[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
% property holds (we believe this is likely true for both
% \citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
% terms of untyped parallel reduction to show the decidability of type
% conversion. Otherwise, to fully model the system, we need to bake
% typing information into our logical relation and we need to redefine
% our logical relation in Kripke-style so we can relax the scoping
% constraint in order to prove adequacy, though a relational model would
% still be unnecessary as long as the $\eta$ laws are baked into the
% typed directed relation. \scw{not sure I understand this last bit}

% Since confluence plays such as key role in our proof, we are uncertain
% how to generalize our technique to systems where confluence does not
% hold, and we will investigate those scenarios in our future work.
% \scw{What systems are you thinking of? Or do you mean more generally
% that this technique only works for languages where the definition of
% equality is based on joinability of parallel reduction?}
